---
title: "Correct First Approach - Differential Expression Analysis"
author: "FGCZ"
date: "`r format(Sys.time(), '%d %B, %Y')`"
running_head: "CorrectFirst DEA"
params:
  wd: ""
  fdr: 0.05
  fc: 0.6
  output_base: !r paste0("PTM_", format(Sys.Date(), "%Y%m%d"))
  utils_script: "src/dea_utils.R"
  phospho_dea_dir: "DEA_setup/DEA_20260109_WUphospho_SHP2_vsn"
  protein_dea_dir: "DEA_setup/DEA_20260109_WUprot_SHP2_vsn"
  annot_file: "DEA_setup/phospho_annot_SHP2_RUX.tsv"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
  bookdown::html_document2:
    toc: true
  pdf_document:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r init, include=FALSE}
# Default parameters - used when running interactively
# Overwritten by params passed to rmarkdown::render()
default_params <- list(
  fdr = 0.05,
  fc = 0.6,
  output_base = paste0("PTM_", format(Sys.Date(), "%Y%m%d")),
  utils_script = "src/dea_utils.R",
  phospho_dea_dir = "DEA_setup/DEA_20260109_WUphospho_SHP2_vsn",
  protein_dea_dir = "DEA_setup/DEA_20260109_WUprot_SHP2_vsn",
  annot_file = "DEA_setup/phospho_annot_SHP2_RUX.tsv"
)

# Use params if available, otherwise use defaults
if (!exists("params") || is.null(params)) {
 params <- default_params
} else {
  # Fill in any missing params with defaults
  for (p in names(default_params)) {
    if (is.null(params[[p]]) || (is.character(params[[p]]) && params[[p]] == "")) {
      params[[p]] <- default_params[[p]]
    }
  }
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Load required libraries
suppressPackageStartupMessages({
  library(prolfqua)
  library(prolfquapp)
  library(prophosqua)
  library(dplyr)
  library(readr)
  library(arrow)
  library(yaml)
  library(writexl)
})
```

# Introduction

This document implements the "Correct First" approach for estimating PTM feature usage. Instead of the standard approach where protein-level corrections are applied after modeling PTM abundances, this approach performs protein normalization **before** modeling.

**Workflow:**
1. Load normalized total proteome abundances
2. Load normalized PTM site abundances
3. Correct PTM abundances by subtracting protein abundances
4. Model the corrected PTM usage values
5. Compute contrasts and evaluate significance
6. Export results

# Data Loading

The three inputs of this analysis are loaded here: the sample annotation that defines the
groups and contrasts, the normalized protein abundances from the total-proteome DEA run,
and the normalized phosphosite abundances from the phospho DEA run. All three come from
runs that already happened; nothing is re-quantified in this report.

## Project Setup

Reads the sample annotation and creates the output directory. The annotation is the
control file of the analysis: its grouping column becomes the model term and its contrast
definitions become the comparisons evaluated further below.

```{r defineParameters}
source(params$utils_script)

wu_id <- "PTM_usage"

# Use output_base param for date-stamped output directories
res_dir <- file.path(params$output_base, "PTM_CF_DPU")
if (!dir.exists(res_dir)) {
  dir.create(res_dir, recursive = TRUE)
}

# Load annotation from direct path
annot <- readr::read_tsv(params$annot_file, show_col_types = FALSE)
knitr::kable(
  head(annot),
  caption = paste0(
    "First rows of the sample annotation driving this CorrectFirst analysis: one row per ",
    "sample, with the grouping columns that define the linear model and the contrasts ",
    "evaluated further below."
  )
)
```

The annotation was read from `` `r basename(params$annot_file)` ``.

## Load Normalized Total Proteome Abundances

Loads the protein-level abundances that serve as the correction baseline. Reverse (decoy)
entries are removed and identifiers canonicalized so that sites and proteins can be
matched in the next step.

```{r loadNormalizedAbundances}
# Load from direct DEA directory path
ldata <- arrow::read_parquet(get_dea_parquet(params$protein_dea_dir))
ldata <- ldata |>
  filter(!grepl("^rev_", protein_Id)) |>
  canonicalize_uniprot_ids()
protein_sample_col <- get_dea_sample_name_column(params$protein_dea_dir)

tot_d <- ldata |>
  dplyr::select(tidyselect::all_of(protein_sample_col), protein_Id,
                normalized_abundance)

n_protein_measurements <- nrow(tot_d)
```

`r format(n_protein_measurements, big.mark = ",")` normalized protein abundance
measurements were loaded.

## Load Normalized Single-Site Abundances

Loads the site-level abundances, the quantity whose usage is being tested.

```{r loadNormalizedSiteAbundances}
# Load from direct DEA directory path
ldata <- arrow::read_parquet(get_dea_parquet(params$phospho_dea_dir))
list <- yaml::read_yaml(get_dea_yaml(params$phospho_dea_dir))
config <- prolfqua::list_to_AnalysisConfiguration(list)
ptm_data <- prolfqua::LFQData$new(ldata, config)
ptm_sample_col <- get_dea_sample_name_column(params$phospho_dea_dir)
n_site_measurements <- nrow(ptm_data$data_long())
```

`r format(n_site_measurements, big.mark = ",")` normalized phosphosite abundance
measurements were loaded.

# Protein Normalization

## Merge and Correct PTM Data

Correct PTM abundances by subtracting protein abundances (on log scale this is equivalent to dividing).

```{r correctData}
sample_join <- stats::setNames(protein_sample_col, ptm_sample_col)
join_columns <- c(sample_join, protein_Id = "protein_Id")

ptm_data$set_data(dplyr::inner_join(
  ptm_data$data_long(), tot_d,
  by = join_columns,
  suffix = c(".site", ".total")
))

ptm_data$set_data(ptm_data$data_long() |>
  dplyr::mutate(ptm_usage = normalized_abundance.site - normalized_abundance.total))

n_merged_measurements <- nrow(ptm_data$data_long())
```

Joining sites to their proteins leaves `r format(n_merged_measurements, big.mark = ",")`
measurements that carry both a site and a protein value and can therefore be corrected;
sites whose protein was not quantified drop out here.

```{r correctDataWrite}
ptm_data$get_config()$set_response("ptm_usage")
wide <- ptm_data$data_wide()
wide_data <- prolfqua::separate_hierarchy(wide$data, ptm_data$get_config())
writexl::write_xlsx(wide_data, file.path(res_dir, "CorrectFirst_intensities_wide.xlsx"))
writexl::write_xlsx(wide$annotation, file.path(res_dir, "CorrectFirst_intensities_file_annotation.xlsx"))
```

## Visualize Corrected PTM Usage

```{r fig1densityCorrected, fig.cap="Distribution of the protein-corrected phosphosite abundances, one curve per sample. The x-axis is the corrected abundance on log2 scale (site abundance minus the abundance of its protein in the same sample, shifted by a constant to keep values positive) and the y-axis its density. Curves that overlay each other indicate comparable correction across samples; a shifted or differently shaped curve marks a sample whose correction behaves differently and that will influence the model."}
ptm_data$get_config()$set_response("ptm_usage")
pl <- ptm_data$get_Plotter()
ptm_data$set_data(ptm_data$data_long() |>
  dplyr::mutate(ptm_usage = ptm_usage + 20))
pl$intensity_distribution_density()

```

```{r pca, fig.cap="Principal component analysis of the protein-corrected phosphosite abundances. Each point is one sample positioned by its first two principal components, coloured by experimental group and labelled with the sample name. Missing values are imputed for this projection only, not for the statistical model. Samples of the same group clustering together indicates that PTM usage separates the conditions; an outlying sample is a candidate explanation for weak contrasts."}
pca_data <- prolfqua::AggregateLimpa$new(
  ptm_data,
  impute_only = TRUE
)$aggregate()
pca_data$get_Plotter()$pca()
```

# Statistical Modeling

The corrected usage values are now treated as an ordinary response: one linear model per
phosphosite, then the contrasts of interest evaluated on those models. Because the
correction already happened, no protein-level adjustment is applied afterwards.

## Fit Linear Model

Fits `ptm_usage ~ G_` per site, that is one group effect per experimental group.

```{r fitModelToCorrected}
strategy_lm <- prolfqua::strategy_lm("ptm_usage ~ G_")
models <- prolfqua::build_model(data = ptm_data, model_strategy = strategy_lm)
n_models <- nrow(models$model_df)

```

One model was fitted per phosphosite, `r format(n_models, big.mark = ",")` in total.

## Define Contrasts

Determines which group comparisons to evaluate. Two annotation styles are accepted: an
explicit one carrying `Contrast` and `ContrastName` columns, or a dataset style carrying
`Group` and `Control`, from which every treatment-versus-control pair is generated.

```{r defineContrasts}
# Support two annotation formats:
# 1. Explicit: has Contrast and ContrastName columns
# 2. Dataset: has Group and Control columns (C=control, T=treatment)

if (all(c("Contrast", "ContrastName") %in% colnames(annot))) {
  # Format 1: Explicit contrasts
  contrasts <- annot$Contrast
  names(contrasts) <- annot$ContrastName
  contrasts <- contrasts[!is.na(contrasts) & nchar(contrasts) > 0]
} else if (all(c("group", "control") %in% tolower(colnames(annot)))) {
  # Format 2: Derive contrasts from Group/Control columns
  # Control='C' means control group, Control='T' means treatment.
  # The column names are matched case-insensitively: annotation files in the wild
  # spell the second column both "Control" and "CONTROL".
  group_col <- colnames(annot)[tolower(colnames(annot)) == "group"][[1]]
  control_col <- colnames(annot)[tolower(colnames(annot)) == "control"][[1]]

  levels_df <- annot |>
    dplyr::select(dplyr::all_of(c(group_col, control_col))) |>
    dplyr::distinct()

  control_groups <- levels_df[[group_col]][toupper(levels_df[[control_col]]) == "C"]
  treatment_groups <- levels_df[[group_col]][toupper(levels_df[[control_col]]) == "T"]

  # Generate contrasts: treatment - control (with G_ prefix for prolfqua)
  contrasts <- character()
  for (trt in sort(treatment_groups)) {
    for (ctrl in sort(control_groups)) {
      contrast_name <- paste0(trt, "_vs_", ctrl)
      contrast_formula <- paste0("G_", trt, " - G_", ctrl)
      contrasts[contrast_name] <- contrast_formula
    }
  }
} else {
  stop("Annotation file must have either (Contrast, ContrastName) or (Group, Control) columns")
}

# An annotation can carry the right columns and still define nothing, for example a
# template whose Group/Control cells are still empty. Fail here rather than fitting
# contrasts that do not exist.
if (length(contrasts) == 0) {
  stop(
    "No contrasts could be derived from ", basename(params$annot_file),
    ": the annotation carries the expected columns but marks no control (C) and ",
    "treatment (T) groups.",
    call. = FALSE
  )
}

knitr::kable(
  data.frame(Contrast_Name = names(contrasts), Contrast = contrasts),
  caption = paste0(
    "Contrasts evaluated on the corrected PTM usage. Contrast_Name is the label used ",
    "throughout the downstream reports, Contrast the linear-model expression it stands for, ",
    "with G_ marking the group coefficients of the fitted model; a positive fold change means ",
    "higher PTM usage in the first named group."
  )
)
```

## Compute Contrasts

Evaluates the contrasts on the fitted models and applies variance moderation, which
stabilises the per-site variance estimates by borrowing information across all sites.

```{r computeTheContrasts}
ctr <- prolfqua::Contrasts$new(models, contrasts)
ctr_df <- ctr$get_contrasts()

ctr <- prolfqua::ContrastsModerated$new(ctr)
ctr_df <- ctr$get_contrasts()

n_site_contrast <- nrow(ctr_df)
```

`r format(n_site_contrast, big.mark = ",")` site-contrast pairs were estimated
(`r format(n_models, big.mark = ",")` sites x `r length(contrasts)` contrasts, minus sites the model could not
estimate for a given contrast).

# Results

## Volcano Plot

Overview of the result: effect size against significance, one panel per contrast.

```{r figVolcanoModelOnly, fig.cap="Differential PTM usage per contrast, from the model fitted on protein-corrected abundances. Each point is one phosphosite; the x-axis is the estimated log2 fold change of corrected usage and the y-axis -log10(FDR) from the moderated t-test, so points to the left and right are sites used less or more, and points high up are the confident ones. Sites imputed rather than measured in a condition are excluded from the exported results."}
ctr_plotter_model <- ctr$get_Plotter()
ctr_plotter_model$volcano()$FDR
```

# Export Results

## Add Site Information from DEA Results

Joins the sequence window, residue position and gene annotation of each site onto the
contrast results, so the exported table is self-contained for the downstream motif and
enrichment reports.

```{r addSiteInfo}
site_info <- get_dea_ptm_site_info(params$phospho_dea_dir) |>
  dplyr::select(-protein_Id)
contrast_site_col <- intersect(c("site", "protein_Id_site"), names(ctr_df))[[1]]
ctr_df_seq_window <- dplyr::left_join(
  ctr_df,
  site_info,
  by = stats::setNames("site", contrast_site_col)
)

```

## Export to Excel

The contrast results are written to a single-sheet Excel workbook, which is the file every
downstream report of this analysis reads. Only site-contrast pairs estimated from measured
values are exported: a pair whose estimate would rest on an imputed value in one of the two
groups is dropped, because a fold change measured against an imputed level is not
interpretable. The table below reports how many pairs each case contributes.

Column names follow the same convention as the DPA and DPU results, so the reports can read
all three interchangeably: `diff.site` is the log2 fold change of protein-corrected PTM
usage, `FDR.site` its adjusted p-value and `statistic.site` the moderated t-statistic.
Alongside these sit the site annotation columns (`protein_Id`, `site`, `posInProtein`,
`modAA`, `SequenceWindow`, `gene_name`, `protein_length`) and `estimate_type`, which records
whether the underlying value was observed or filled in by limit-of-detection imputation.

```{r exportResults}
# Rename columns to match DPA convention (diff.site, FDR.site, statistic.site)
ctr_df <- ctr_df_seq_window |>
  dplyr::rename(diff.site = diff, FDR.site = FDR, statistic.site = statistic) |>
  dplyr::mutate(
    modelName = ifelse(
      modelName %in% c("WaldTest", "WaldTest_moderated"),
      "Linear_Model_Moderated",
      "MissingInOneCondition"
    )
  )

# Count both cases before filtering, so the report can state what was dropped.
model_counts <- ctr_df |>
  dplyr::count(modelName, name = "Site_contrast_pairs")
n_before <- nrow(ctr_df)

# Remove contrasts computed with imputation
ctr_df <- ctr_df |> dplyr::filter(modelName != "MissingInOneCondition")

# The filter is allowed to drop rows; what must not happen is dropping everything.
stopifnot(
  "no site-contrast pair survived the imputation filter" = nrow(ctr_df) > 0
)


output_file <- file.path(res_dir, paste0("CorrectFirst_", wu_id, "_results.xlsx"))
writexl::write_xlsx(
  list(results = ctr_df),
  path = output_file
)

knitr::kable(
  model_counts,
  caption = paste0(
    "How the site-contrast pairs were estimated, counted before the imputation filter. ",
    "Linear_Model_Moderated pairs come from the moderated linear model on measured values ",
    "and are exported; MissingInOneCondition pairs would rest on an imputed value in one of ",
    "the two groups and are dropped. Exported: ",
    format(nrow(ctr_df), big.mark = ","), " of ", format(n_before, big.mark = ","),
    " site-contrast pairs, over ", dplyr::n_distinct(ctr_df$contrast), " contrasts."
  )
)
```

The workbook is named `` `r basename(output_file)` `` and its single sheet is called
`results`. It holds `r nrow(ctr_df)` rows and `r ncol(ctr_df)` columns.

## Output Location

Every file produced by this analysis is written to `` `r res_dir` ``:

| File | Content |
|:-----|:--------|
| `` `r basename(output_file)` `` | the contrast results exported above, sheet `results`. This is the file the N-to-C, seqlogo and enrichment reports read. |
| `CorrectFirst_intensities_wide.xlsx` | the protein-corrected PTM usage matrix, one column per sample, as it entered the linear model. |
| `CorrectFirst_intensities_file_annotation.xlsx` | the sample annotation belonging to that matrix. |
| `Analysis_CorrectFirst_DEA.html` | this report. |

# Summary

This "Correct First" approach:

1. **Normalizes PTM abundances by protein abundances** before statistical modeling
2. **Models the corrected PTM usage** values directly
3. **Computes contrasts** on the protein-normalized PTM values

This differs from the "correct last" approach (DPU/MSstatsPTM method) where PTM and protein abundances are modeled separately, then protein-level corrections are applied at the contrast level.

# Session Info

```{r sessionInfo}
sessionInfo()
```
