---
title: "DPA and DPU Integration Analysis"
author:
  - FGCZ
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
params:
  output_base: !r paste0("PTM_", format(Sys.Date(), "%Y%m%d"))
  utils_script: "src/dea_utils.R"
  phospho_dea_dir: "DEA_setup/DEA_20260109_WUphospho_SHP2_vsn"
  protein_dea_dir: "DEA_setup/DEA_20260109_WUprot_SHP2_vsn"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r setupglobal, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

suppressPackageStartupMessages({
  library(prolfquapp)
  library(prophosqua)
  library(dplyr)
})

```

# Introduction

This analysis integrates total proteome and phosphoproteome DEA results to compute:

- **DPA** (Differential PTM Abundance): Raw phosphosite signal changes
- **DPU** (Differential PTM Usage): Protein-normalized changes (site - protein) reflecting modification stoichiometry

# Setup

```{r b1specifyinputs}
library(prophosqua)

wu_id <- "CustomPhosphoAnalysis"
fgcz_project <- "PTM_analysis"
oid_fgcz <- "fgcz_project"
datetoday <- format(Sys.Date(), "%Y%m%d")
project_dir <- "."

# Use output_base param for date-stamped output directories
res_dir_dpa <- file.path(params$output_base, "PTM_DPA")
res_dir_dpu <- file.path(params$output_base, "PTM_DPU")
```

```{r b2createoutputdir}
descri <- wu_id
if (!dir.exists(file.path(project_dir, res_dir_dpa))) {
  dir.create(file.path(project_dir, res_dir_dpa), recursive = TRUE)
}
if (!dir.exists(file.path(project_dir, res_dir_dpu))) {
  dir.create(file.path(project_dir, res_dir_dpu), recursive = TRUE)
}
```

# Load DEA Results

```{r definePaths}
source(params$utils_script)

# Get Excel files from direct DEA directory paths
tot_file <- get_dea_xlsx(params$protein_dea_dir)
ptm_file <- get_dea_xlsx(params$phospho_dea_dir)

stopifnot(file.exists(tot_file))
stopifnot(file.exists(ptm_file))
```

```{r b4loaddata}
required_cols <- c("protein_Id", "protein_length", "contrast")
tot_res <- prophosqua::load_and_preprocess_data(tot_file, required_cols)
tot_res <- prophosqua::filter_contaminants(tot_res)
tot_res <- canonicalize_uniprot_ids(tot_res)
phospho_res <- prophosqua::load_and_preprocess_data(ptm_file, required_cols)
phospho_res <- prophosqua::filter_contaminants(phospho_res)

site_info <- get_dea_ptm_site_info(params$phospho_dea_dir) |>
  dplyr::select(site, posInProtein, modAA, SequenceWindow)
phospho_site_col <- intersect(c("site", "protein_Id_site"), names(phospho_res))[[1]]
phospho_res <- dplyr::left_join(
  phospho_res,
  site_info,
  by = stats::setNames("site", phospho_site_col)
)
```

# DPA: Integrate PTM and Protein Data

```{r b5combinedata}
join_column <- c(
  "protein_Id" = "protein_Id",
  "contrast",
  "description",
  "protein_length"
)
suffix_a <- ".site"
suffix_b <- ".protein"
phospho_res <- phospho_res |> filter(!is.na(FDR))

combined_site_prot <- dplyr::left_join(
  phospho_res,
  tot_res,
  by = join_column,
  suffix = c(suffix_a, suffix_b)
)
```

## Match Rate

```{r b5matchrate}
match_rates <- combined_site_prot |>
  group_by(contrast) |>
  summarize(
    total_sites = n(),
    matched_sites = sum(!is.na(diff.protein)),
    match_rate = round(matched_sites / total_sites * 100, 1)
  )

knitr::kable(
  match_rates,
  col.names = c("Contrast", "Total Sites", "Matched Sites", "Match Rate (%)"),
  caption = paste0(
    "How many phosphosites could be paired with a protein-level result, per contrast. Total ",
    "Sites counts the sites tested in the phospho DEA, Matched Sites those whose protein was ",
    "also quantified in the total proteome DEA, and Match Rate their ratio. Only matched ",
    "sites can be corrected for protein abundance, so unmatched sites carry a DPA result but ",
    "no DPU value."
  )
)
```

# DPU: Compute Differential PTM Usage

DPU calculates protein-normalized PTM changes:

- Effect size: $\Delta_I = \log_2FC_{PTM} - \log_2FC_{Protein}$
- Standard error: $SE_I = \sqrt{SE_{PTM}^2 + SE_{Protein}^2}$

```{r b8computeptmusage}
combined_test_diff <- prophosqua::test_diff(phospho_res, tot_res, join_column = join_column)

# Add 'site' alias for compatibility with template (new format uses protein_Id_site)
if (!"site" %in% colnames(combined_test_diff) && "protein_Id_site" %in% colnames(combined_test_diff)) {
  combined_test_diff <- combined_test_diff |> dplyr::mutate(site = protein_Id_site)
}
```

# Export Results

```{r storeptmusagexlsx}
writexl::write_xlsx(
  list(combinedStats = combined_test_diff),
  path = file.path(project_dir, res_dir_dpu, "Result_DPU.xlsx")
)

writexl::write_xlsx(
  list(combinedSiteProteinData = combined_site_prot),
  path = file.path(project_dir, res_dir_dpa, "Result_DPA.xlsx")
)

# Save intermediate data for DPU overview rendering (separate step)
saveRDS(
  combined_test_diff,
  file = file.path(project_dir, res_dir_dpu, "combined_test_diff.rds")
)
```

# Output Files

- `PTM_DPA/Result_DPA.xlsx` - Differential PTM Abundance (raw PTM changes)
- `PTM_DPU/Result_DPU.xlsx` - Differential PTM Usage (protein-normalized changes)

# Session Info

```{r sessioninfo}
sessionInfo()
```
